Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model

### 2025-09-15 paper/paper.qmd
acronyms markdown and feedback

---
title: Paper
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
---

@@

```{r load-data, include=FALSE}
# print(file.exists(here::here("scripts", "calibrate.R")))
# source(here::here("scripts", "calibrate.R"))
source(file.path("..", "scripts", "calibrate.R"), chdir = TRUE)
# load(file.path("..", "out", "calibrate.RData"))
```

Transmission of potentially pathogenic bacteria under antimicrobial selective pressure has important implications for control of hospital-acquired infection.
Examples of such pathogens include *Clostridioides difficile* infection (CDI), vancomycin-resistant *Enterococcus* (VRE), extended-spectrum beta-lactamase-producing (ESBL) Enterobacterales, and carbapenem-resistant Enterobacterales (CPE).
Infection with these pathogens can lead to increased morbidity and mortality -- for example, CPE bloodstream infection is associated with an increased risk of in-hospital mortality (risk difference of 0.25) compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
_C. difficile_ infection is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
Furthermore, the therapeutic antimicrobial options available for VRE, ESBL, and CPE are limited, leading to increased use of WHO reserve agents, which further exacerbates the problem of antimicrobial resistance [@organization2019WHOAWaRe2019].

Infection prevention and control (IPC) strategies are essential to prevent the spread of these pathogens within a healthcare setting, and include measures such as hand hygiene, environmental cleaning, isolation of infected/colonised patients, and more [@worldhealthorganizationGlobalReportInfection2022].
In many healthcare settings, IPC strategies and operational practices are led by a multi-disciplinary IPC team.
A key role of IPC teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected/colonised patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such rates of antimicrobial usage, and to structural differences such as hospital layout.

In practice, the approach to managing outbreaks is often reactive.
Once an outbreak is identified and declared, the IPC team implements a set of enhanced infection control measures to attempt to attempt to halt transmission.
The number of cases of infection or colonisation is monitored over a time frame dependent on the pathogen, setting, and scale of the outbreak.
If the measures are successful in controlling the spread, then they are gradually relaxed once rates return to baseline.
If they are not, then further measures are implemented, or the existing measures are intensified, until control is achieved.
A inherent limitation of this approach is that decisions are made based on observed data, and the impact of interventions cannot be quantified until after a change in the rate of infection has been observed, by which time further preventable nosocomial infections may have occurred.

Computer simulation approaches such as agent-based modelling (ABM) can be used to simulate the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact. Agent-based modelling has been used to model various infectious diseases, such as the transmission of important nosocomial pathogens such as methicillin-resistant _Staphylococcus aureus_ (MRSA) within a healthcare setting [@jaramilloAgentBasedModel2015], and the spread of SARS-CoV-2 within a community population [@kerrCovasimAgentbasedModel2021]. An important advantage of ABM is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019]. This can be particularly useful in the preparedness for, and management of, outbreaks of healthcare-associated infections, where the impact of interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019]. "What if?" testing allows for different scenarios or bundles of interventions can be simulated to determine their impact.

In this study, we describe an agent-based model that simulates the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. The aim of the model is to guide infection control strategies and interventions, particularly during outbreak periods. The parameters are estimated using data from a CDI outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although the model is designed to be adaptable to other related antimicrobial-resistant pathogens such as VRE, ESBL, and CPE.

@@

An agent-based model (ABM) was implemented using NetLogo [@wilenskyu.NetLogo1999]. Briefly, the simulation started by building a virtual hospital with a configurable high-level structure, defined by its number of wards, beds per ward, and proportion of bedspaces sharing toilet facilities. Next, the hospital was populated with patients, each having an infected/colonised or non-infected/colonised state, defined as:

@@

-   $x_p(t) = 0$ if patient $p$ is not infected/colonised at time $t$,
-   $x_p(t) = 1$ if patient $p$ is infected/colonised at time $t$.

It was assumed that infection/colonisation was not reversible during the simulation, i.e., that infected/colonised patients remained so until discharge. At each time increment, the probability of a non-colonised/infected patient acquiring infection (i.e., transitioning from a non-infected/colonised to infected/colonised state) depended on: 1) a baseline risk of *de novo* infection/colonisation:

@@

-   $R_i(t)$ is the risk of patient $i$ acquiring infection/colonisation at time $t$,

@@

The agent-based model conducted Bernoulli trials to determine if a patient acquired infection/colonisation at each time step $t$:

@@

where $X_\text{de novo}$ = 1 if patient acquires infection/colonisation *de novo* at time $t$, and 0 otherwise.

@@

where $X_\text{toilet}$ = 1 if patient acquires infection/colonisation from a contaminated toilet at time $t$, and 0 otherwise.

@@

where $X_i(t)$ = 1 if patient $i$ acquires infection/colonisation at time $t$, and 0 otherwise.

@@

To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures.

@@

The outputs of the model included: 1) number of patients admitted; 2) number of patients with community-acquired colonisation/infection; 3) number of patients with hospital-acquired (nosocomial) colonisation/infection. The simulation was run with each unit of time corresponding to one day; therefore, the model's outputs were for each simulated day.

@@

Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgadoAnalysisOutbreakClostridium2009]. Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].

@@

Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnakerApproximateBayesianComputation2013]. Priors were determined from domain knowledge or extracted from existing literature. A summary of the priors used is available in @tbl-priors.

Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$. Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets. The number of daily cases of nosocomial CDI were extracted from the model outputs, and converted to a time series of nosocomial CDI per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$. Results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.

@@

Otherwise, $S_i$ was rejected. The parameter samples that generated the accepted simulations were used to produce a posterior distribution. $\epsilon$ was set so that the acceptance rate was approximately 2.5%.

A two-step approach to calibration was taken. Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that simulated the baseline rate. Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009]. To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation. The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency. To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded. At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures. The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters. Again, only samples from the latin hypercube that satisfied these requirements were used.

@@

Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection/colonisation were calculated. For the samples used for ABC, `r results_outbreak_control_n_seeds` different random number generator seeds were used;
for the posterior predictive checks, `r uniform_sims_n_seeds` seeds were used for the uniform distribution,
`r mean_sims_n_seeds` seeds were used for the mean of the posterior distribution, and
`r posterior_predictive_n_seeds` seeds were used for the samples with replacement from the posterior distribution.

One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method. Since the agent-based model was stochastic, all simulations were aggregated

@@

To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019]. Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) proportion of patients redistributed to a single-occupancy bed after colonisation/infection; 3) toilet cleaning rate; and 4) toilet cleaning effect. To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection ABC. Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value. The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period. For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.

@@

An agent-based model that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced. After calibrating the agent-based model using approximate Bayesian computation (ABC), three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control. A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection ABC) is shown in @tbl-posteriors. The densities of the posterior parameter distributions are shown in @fig-posterior-densities. In line with the prior bounds described in @sec-outbreak-generation, the posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).

@@

```{r fig-posterior-densities, include=TRUE}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| warning: false
#| label: fig-posterior-densities
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
#| fig-subcap:
#|   - "General parameters"
#|   - "Toilet-related parameters"
#| layout-ncol: 1
abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))

@@

Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check). In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\]. During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-posterior-predictive-check).

```{r fig-abc-rates-retained-posteriors, include=TRUE}
#| echo: false
#| label: fig-abc-rates-retained-posteriors
#| fig.cap: "Simulated rates of _C. difficile_ infection generated from the simulations retained after applying the rejection approximate Bayesian computation algorithm. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
#| fig.width: 18
#| fig.height: 8
ggplot(long_observed_ss, aes(x = x, y = rates)) + geom_line(color = 'red') +
  geom_line(data = sim_mean, aes(x = x, y = m)) +
  geom_line(data = sim_mean, aes(x = x, y = q_high), linetype='dashed') +
  geom_line(data = sim_mean, aes(x = x, y = q_low), linetype='dashed') + 
  theme(text=element_text(size=24)) +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

@@

```{r fig-consts-sims-mean-sd, include=TRUE}
#| echo: false
#| fig-cap: "Simulated rates of _C. difficile_ infection generated by using the posterior parameter distributions. Input parameter samples were generated by sampling the uniform distribution $U_[\\mu - \\sigma, \\mu + \\sigma]$, where $\\mu$ and $\\sigma$ are the mean and standard deviation of the parameter's posterior distribution respectively. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
# plot but with mean and sd instead rather than individual
uniform_sims_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim) %>%
  summarise(m = median(rate),
            q_high = quantile(rate, 0.75),
            q_low = quantile(rate, 0.25),
            .groups = "keep") %>%
  ggplot() +
  geom_line(aes(x = date_sim, y = m)) +
  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
  guides(color="none")
```

>>>>

---
title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: ICR
    longname: interquartile range
---

@@

```{r load-data, include=FALSE, eval=FALSE}
# This runs the calibrate scripts and loads objects into memory.
# print(file.exists(here::here("scripts", "calibrate.R")))
# source(here::here("scripts", "calibrate.R"))
source(file.path("..", "scripts", "calibrate.R"), chdir = TRUE)
# load(file.path("..", "out", "calibrate.RData"))
```

```{r load-pre-processed-data, include=FALSE}
# Alternatively, just load previously dumped .RData from a run of calibrate.R
load(file.path("..", "out", "calibrate.RData"))
```

The transmission of pathogenic bacteria in hospitals results in excessive morbidity and mortality.
Relevant examples include *Clostridioides difficile*, \acr{VRE}, \acr{ESBL}-producing Enterobacerales, and \acr{CPE}.
Drug resistance further compounds the problem --- for example, \acr{CPE} bloodstream infection is associated with an increased risk of in-hospital mortality (risk difference of 0.25) compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
\acr{CDI} is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
Furthermore, the therapeutic antimicrobial options available for \acr{VRE}, \acr{ESBL}, and \acr{CPE} are limited, leading to increased use of \acr{WHO} reserve agents, which further exacerbates the problem of antimicrobial resistance [@organization2019WHOAWaRe2019].

\acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings. Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected/colonised patients [@worldhealthorganizationGlobalReportInfection2022].
A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected/colonised patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such rates of antimicrobial usage, and to structural differences such as hospital layout.

In practice, the approach to managing outbreaks is often reactive.
The number of cases of infection is monitored over a time frame dependent on the pathogen, setting, and scale of the outbreak.
If the measures are successful in controlling the spread, then they are gradually relaxed once rates return to baseline.
If they are not, then further measures are implemented, or the existing measures are intensified, until control is achieved.
A inherent limitation of this approach is that decisions are made based on observed data, and the impact of interventions cannot be quantified until after a change in the rate of infection has been observed, by which time further preventable nosocomial infections may have occurred.

The use of computer simulation techniques such as \acr{ABM} can be used to model the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact. Agent-based modelling has been used to describe the transmission of \acr{MRSA} [@jaramilloAgentBasedModel2015] and \acr{SARS-CoV-2} [@kerrCovasimAgentbasedModel2021]. An advantage of \acr{ABM} over other simulation approaches is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019]. This can be particularly useful for both planning and active management of outbreaks, where the impact of individual interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019].

Here, we describe an agent-based model that describes the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.

@@

An \acr{ABM} was implemented using NetLogo [@wilenskyu.NetLogo1999]. Briefly, the simulation started by building a virtual hospital with a configurable high-level structure, defined by its number of wards, beds per ward, and proportion of bedspaces sharing toilet facilities. Next, the hospital was populated with patients, each having an infected or non-infected state, defined as:

@@

-   $x_p(t) = 0$ if patient $p$ is not infected at time $t$,
-   $x_p(t) = 1$ if patient $p$ is infected at time $t$.

Since the simulation's target diseases were associated with long term colonisation, and risk of onward transmission, after the initial infection event, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained colonised until discharge. At each time increment, the probability of a non-colonised/infected patient acquiring infection (i.e., transitioning from a non-infected/colonised to infected/colonised state) depended on: 1) a baseline risk of *de novo* infection/colonisation:

@@

-   $R_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,

@@

The agent-based model conducted Bernoulli trials to determine if a patient became infected at each time step $t$:

@@

where $X_\text{de novo}$ = 1 if patient becomes infected *de novo* at time $t$, and 0 otherwise.

@@

where $X_\text{toilet}$ = 1 if patient becomes infected from a contaminated toilet at time $t$, and 0 otherwise.

@@

where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.

@@

To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.

@@

The outputs of the model included: 1) number of patients admitted to the hospital; 2) number of patients with community-acquired infection; 3) number of patients with hospital-acquired (nosocomial) infection. The simulation was run with each unit of time corresponding to one day; therefore, the model's outputs were for each simulated day.

@@

Data from a hospital \acr{CDI} outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgadoAnalysisOutbreakClostridium2009]. Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].

@@

Rejection sampling \acr{ABC} was employed to fit the model to the outbreak data [@sunnakerApproximateBayesianComputation2013]. Priors were extracted from existing literature; flat priors when data were unavailable. A summary of the priors used is available in @tbl-priors. The proportion of toilets that were shared between multiple patients was set to 60%.

Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$. Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets. The number of daily cases of nosocomial \acr{CDI} were extracted from the model outputs, and converted to a time series of nosocomial \acr{CDI} per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$. Since each simulation starts with an empty hospital that is gradually populated with patient admissions, results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.

@@

Otherwise, $S_i$ was rejected. The parameter samples that generated the accepted simulations were used to produce a posterior distribution. $\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.

A two-step approach to calibration was taken. Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that recapitulated the baseline rate. Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009]. To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation. The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency. To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded. At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures. The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters. Again, only samples from the latin hypercube that satisfied these requirements were used.

@@

Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection were calculated. For the samples used for ABC, `r results_outbreak_control_n_seeds` different random number generator seeds were used;
for the posterior predictive checks, `r uniform_sims_n_seeds` seeds were used for the uniform distribution,
`r mean_sims_n_seeds` seeds were used for the mean of the posterior distribution, and
`r posterior_predictive_n_seeds` seeds were used for the samples with replacement from the posterior distribution.

One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method.

@@

To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019]. Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) proportion of patients redistributed to a single-occupancy bed after infection; 3) toilet cleaning rate; and 4) toilet cleaning effect. To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value. The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period. For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.

@@

An \acr{ABM} that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced. After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control. A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors. The densities of the posterior parameter distributions are shown in @fig-posterior-densities. In line with the prior bounds described in @sec-outbreak-generation, the posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).

@@

```{r fig-posterior-densities, include=TRUE}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| warning: false
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
#| fig-subcap:
#|   - "General parameters"
#|   - "Toilet-related parameters"
#| layout-ncol: 1
abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))

@@

Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check). In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[\acr{ICR} range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\]. During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[\acr{IQR} `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[\acr{IQR} 3.32--4.41\] (see @fig-posterior-predictive-check).

```{r fig-abc-rates-retained-posteriors, include=TRUE}
#| echo: false
#| fig.cap: "Simulated rates of _C. difficile_ infection generated from the simulations retained after applying the rejection approximate Bayesian computation algorithm. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
#| fig.width: 18
#| fig.height: 8
ggplot(long_observed_ss, aes(x = x, y = rates)) + geom_line(color = 'red') +
  geom_line(data = sim_mean, aes(x = x, y = m)) +
  geom_line(data = sim_mean, aes(x = x, y = q_high), linetype='dashed') +
  geom_line(data = sim_mean, aes(x = x, y = q_low), linetype='dashed') + 
  theme(text=element_text(size=24)) +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

@@

```{r fig-consts-sims-mean-sd, include=TRUE}
#| echo: false
#| fig-cap: "Simulated rates of _C. difficile_ infection generated by using the posterior parameter distributions. Input parameter samples were generated by sampling the uniform distribution $U_[\\mu - \\sigma, \\mu + \\sigma]$, where $\\mu$ and $\\sigma$ are the mean and standard deviation of the parameter's posterior distribution respectively. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
# plot but with mean and sd instead rather than individual
uniform_sims_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim) %>%
  summarise(m = median(rate),
            q_high = quantile(rate, 0.75),
            q_low = quantile(rate, 0.25),
            .groups = "keep") %>%
  ggplot() +
  geom_line(aes(x = date_sim, y = m)) +
  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
  guides(color="none") +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```
